HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN

### 2020-04-08 cmc-rnaseq-de/Differential-Expression_Tissue.Rmd
add rRNARate to covars to adjust

---
title: "Differential Expression"
author: "Kelsey Montgomery, Gabriel Hoffman, Thaneer Perumal"
date: "`r date()`"
output: html_document
---

@@

knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/Differential-Expression.Rmd",
                                 parentId ="syn21887806",
                                 folderName = "Differential Expression - Dx.Tissue")
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn21887806';
activityName = 'Differential Expression';
activityDescription = 'Differential Expression - Dx.Tissue';

thisFileName <- 'Differential-Expression.Rmd'

@@

# Set covariates to adjust for from model selection
adjust.covars <- c("Dx.Tissue", "RIN", "IntragenicRate")
```

### Normalisation - Dx.Tissue
1. Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
2. Individual_ID is chosen as random effect

```{r iterative.normalisation, results='asis'}
writeLines(paste('Using following covariates in the model:',
                 paste(adjust.covars, collapse=', '),
                 'as fixed effects and Individual_ID is chosen as random effect'))

# Post adjusted formula
random_effect <- "Individual_ID"
formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
  glue::glue(" + (1|{random_effect})")

# Estimate voom weights with DREAM
geneExpr = DGEList(NEW.COUNTS)
geneExpr = edgeR::calcNormFactors(geneExpr)

VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
                                                               formula = formula,
                                                               data = COVARIATES,
                                                               save.plot = TRUE)

# How to plot a voom object if save.plot=TRUE
# Set plot generic to voom result EList
#' @export
setMethod("plot", signature(x="EList", y="missing"),
  function(x,  y, ...){
    if( is.null(x$voom.xy) || is.null(x$voom.line)){
        stop("x does not contain the trend information.\nvoom() must be run with save.plot=TRUE")
    }
    # points
    df = data.frame(x = x$voom.xy$x,
                    y = x$voom.xy$y)
    # trend line
    df.line = data.frame(x = x$voom.line$x,
                        y = x$voom.line$y)
    ggplot(df, aes(x,y)) + geom_point(size=.1) + theme_bw(15) + 
      theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) + 
      geom_line(data=df.line, aes(x,y), color="red") + xlab(bquote(log[2](count + 0.5))) + 
      ylab(expression( sqrt("standard deviation"))) + ggtitle("Voom: Mean-variance trend") + 
      ylim(0, max(df$y))
  }
)

plot(VOOM.GENE_EXPRESSION)

# Fit linear model using new weights and new design
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION
ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
                                        formula = formula,
                                        data = COVARIATES,
                                        computeResiduals = TRUE)

col_type <- dplyr::select(COVARIATES, c(adjust.covars, "Individual_ID")) %>%
  dplyr::summarise_all(class) %>% 
  tidyr::gather(variable, class)

# Model categorical variables as random to report variance fractions
formula_for_variance_fractions = sapply(1:length(col_type$class), function(i){
  switch(col_type$class[i], 
    "factor" = paste0('(1|', col_type$variable[i], ')'),
    "numeric" = col_type$variable[i])
  })

formula_for_variance_fractions <- glue::glue("~ ", glue::glue_collapse(formula_for_variance_fractions, sep = " + "))

violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
                                                         formula = formula_for_variance_fractions,
                                                         data = COVARIATES)

plotVarPart(sortCols(violin_plot))

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```

### Tissue Correlation
```{r tissue_correlation}
hasReps = rownameToFirstColumn(COVARIATES, "SampleID") %>%
  dplyr::select(Individual_ID) %>% 
  table

df_stg = rownameToFirstColumn(COVARIATES, "SampleID") %>% 
  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
  filter(grepl("STG", Dx.Tissue)) %>% 
  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
df_acc = rownameToFirstColumn(COVARIATES, "SampleID") %>%
  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
  filter(grepl("ACC", Dx.Tissue)) %>% 
  dplyr::select(SampleID, Individual_ID, Dx.Tissue)

df_merge = merge(df_stg, df_acc, by="Individual_ID")

res_std <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id")

corRes = sapply(res_std$ensembl_gene_id, function(ensGene){

  x = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.x,drop=FALSE]
  y = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.y,drop=FALSE]

  cor(t(x)[,1], t(y)[,1], use="pairwise.complete.obs")
})

hist(corRes, xlab="Correlation", main="Correlation between ACC and STG sample from the sample individual for each gene")

pca = prcomp(t(VOOM.GENE_EXPRESSION$E))

df_pca = merge(pca$x, as.data.frame(rownameToFirstColumn(COVARIATES, "SampleID")), 
               by.x="row.names", 
               by.y="SampleID" ) %>% 
  tidyr::separate(Dx.Tissue, into = c("Dx", "Tissue"), sep = "_") %>% 
  dplyr::mutate(Institution = "NIMH-HBCC")

ggplot(df_pca, aes(PC1, PC2, color=Tissue, shape=Institution)) + geom_point() + theme_bw(17) + theme(aspect.ratio=1)
```

@@

# Code
CODE <- Folder(name = "Differential Expression - Dx.Tissue", parentId = parentId)
CODE <- synStore(CODE)

# Store residual gene expression for network analysis
RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_STG_Dx.Tissue_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('CMC_ACC_STG_Dx.Tissue_netResidualExpression.tsv', 
                name = 'Dx.Tissue Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(nEXP_OBJ, annotations = all.annotations)

# Store differential expression results
write.table(DE, file = 'CMC_ACC_STG_Dx.Tissue_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_ACC_STG_Dx.Tissue_DiffExpression.tsv', 
                name = 'Differential Expression Results (Dx.Tissue)', 
                parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(DEXP_OBJ, annotations = all.annotations)

# Store adjusted fit linear mixed model object as RDS - Dx.Tissue
saveRDS(ADJUSTED.FIT, file = 'CMC_ACC_STG_Dx.Tissue_dream-output.rds')
ADJ_OBJ = File('CMC_ACC_STG_Dx.Tissue_dream-output.rds',
               name = 'Dream Output object - Dx.Tissue', 
               parentId = CODE$properties$id)
ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
                   executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
```

>>>>

###Draft

---
title: "Differential Expression - Tissue"
author: "Kelsey Montgomery, Gabriel Hoffman, Thaneer Perumal"
date: "`r date()`"
output: html_document
---

@@

knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/Differential-Expression.Rmd",
                                 parentId ="syn21887806",
                                 folderName = "Differential Expression - Tissue")
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn21887806';
activityName = 'Differential Expression';
activityDescription = 'Differential Expression - Tissue';

thisFileName <- 'Differential-Expression_Tissue.Rmd'

@@

# Set covariates to adjust for from model selection
adjust.covars <- c("Dx.Tissue", "RIN", "rRNARate", "IntragenicRate")
```

### Normalisation - Tissue
1. Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and reported gender)
2. Individual_ID is chosen as random effect
```{r iterative.normalisation-tissue, results='asis'}
###we might not need this adjustment
COVARIATES <- tidyr::separate(COVARIATES, Dx.Tissue.Gender, c("Dx", "Tissue", "Reported_Gender")) %>% 
  dplyr::mutate(Dx = factor(Dx, levels = c("SCZ", "Control", "Other")),
                Tissue = factor(Tissue), 
                Reported_Gender = factor(Reported_Gender))
adjust.covars[1] <- "Dx"
adjust.covars <- c(adjust.covars, "Tissue", "Reported_Gender")
writeLines(paste('Using following covariates in the model:',
                 paste(adjust.covars, collapse=', '),
                 'as fixed effects and Individual_ID is chosen as random effect'))
# Post adjusted formula
random_effect <- "Individual_ID"
formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
  glue::glue(" + (1|{random_effect})")
# Estimate voom weights with DREAM
geneExpr = DGEList(NEW.COUNTS)
geneExpr = edgeR::calcNormFactors(geneExpr)
VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
                                                               formula = formula,
                                                               data = COVARIATES,
                                                               save.plot = TRUE)
plot(VOOM.GENE_EXPRESSION)
# Fit linear model using new weights and new design
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
Tissue_ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
                                        formula = formula,
                                        data = COVARIATES,
                                        computeResiduals = TRUE)
col_type <- dplyr::select(COVARIATES, c(adjust.covars, "Individual_ID")) %>%
  dplyr::summarise_all(class) %>% 
  tidyr::gather(variable, class)
# Model categorical variables as random to report variance fractions
formula_for_variance_fractions = sapply(1:length(col_type$class), function(i){
  switch(col_type$class[i], 
    "factor" = paste0('(1|', col_type$variable[i], ')'),
    "numeric" = col_type$variable[i])
  })
formula_for_variance_fractions <- glue::glue("~ ", glue::glue_collapse(formula_for_variance_fractions, sep = " + "))
violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
                                                         formula = formula_for_variance_fractions,
                                                         data = COVARIATES)
plotVarPart(sortCols(violin_plot))
# Residuals after normalisation
Tissue_RESIDUAL.GENE_EXPRESSION = residuals(Tissue_ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(Tissue_RESIDUAL.GENE_EXPRESSION, COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

### Sanity check
```{r residualSigCovars.manual-tissue, fig.width=12, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```

### Differential expression analysis (ACC - DLPFC)
Genes that are differentially expressed at an FDR <= 0.05 are:
```{r set-formula-tissue}
formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
  glue::glue(" + (1|{random_effect})")
```

```{r diffExp-tissue, fig.height=10, fig.width=15}
L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("TissueDLPFC"))
L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("DxSCZ", "DxControl"))
L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Reported_GenderMale"))
L = cbind(L1, L2, L3)
contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
              glue::glue_collapse(names(L2[L2 != 0]), "-"),
              glue::glue_collapse(names(L3[L3 != 0]), "-")
)
# Visualize contrast matrix
plotContrasts(L) 
# Fit contrast
FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
                                     formula = formula,
                                     data = COVARIATES,
                                     L = L)
# Get differential expression
DE_3 = lapply(glue::glue("L{1:3}"), function(i, FIT){
  topTable(FIT, coef=i, number = Inf) %>%
    rownameToFirstColumn('gene_id') %>% 
    left_join(backgroundGenes)
}, FIT.CONTR) 
names(DE_3) <- contrast
DE_3 = DE_3%>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>% 
  dplyr::mutate(Comparison = ifelse(Comparison %in% "Reported_GenderMale", "Female-Male", Comparison),
                Comparison = ifelse(Comparison %in% "DxSCZ-DxControl", "SCZ-Control", Comparison),
                Comparison = ifelse(Comparison %in% "TissueDLPFC", "ACC-DLPFC", Comparison)) %>% 
  left_join(Ensemble2HGNC) %>%
  left_join(gene_lengths %>% rownameToFirstColumn('gene_id'))
DE_3$Direction[DE_3$adj.P.Val > 0.05 | abs(DE_3$logFC) < log2(1.2)] = 'NONE'
tmp = DE_3 %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
tmp1 = DE_3 %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
kable(full_join(tmp,tmp1))
p = ggplot(DE_3, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme_bw() %+replace% theme(legend.position = 'top')
p = p + facet_grid(Comparison~., scales = 'fixed')
p
```

@@

# Code
CODE <- Folder(name = "Differential Expression - Tissue", parentId = parentId)
CODE <- synStore(CODE)

# Store residual gene expression for network analysis
RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_STG_Tissue_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('CMC_ACC_STG_Tissue_netResidualExpression.tsv', 
                name = 'Tissue Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(nEXP_OBJ, annotations = all.annotations)

# Store differential expression results
write.table(DE, file = 'CMC_ACC_STG_Tissue_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_ACC_STG_Tissue_DiffExpression.tsv', 
                name = 'Differential Expression Results (Tissue)', 
                parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(DEXP_OBJ, annotations = all.annotations)

# Store adjusted fit linear mixed model object as RDS - Tissue
saveRDS(ADJUSTED.FIT, file = 'CMC_ACC_STG_Tissue_dream-output.rds')
ADJ_OBJ = File('CMC_ACC_STG_Tissue_dream-output.rds',
               name = 'Dream Output object - Tissue', 
               parentId = CODE$properties$id)
ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
                   executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
```
